A Higher-Order Approach to Explicit Contexts in LF

### 2015-03-08 paper/paper.tex
hardening proof of theorem 6.1
Add \dlabel macro for labelling derivations
proved sigma elims; will circle back after I try it in twelf

\newcommand\weaken[3]{\mathsf{wk}_{\InputMode{#1}}\,\InputMode{#2}\leadsto\OutputMode{#3}}
\newcommand\qisset[1]{#1\;`\mathit{set}}
\newcommand\qeqset[2]{#1=#2\;`\mathit{set}}
\newcommand\qmem[2]{#1\;`\!\!\in#2}
\newcommand\qeqmem[3]{#1=#2\;`\!\!\in#3}

@@

@@

The next thing to do is to specify how each of these judgement(-codes) behaves
operationally (i.e. by substituting its outermost variable); we will do so by
extending the LF signature with a new algorithmic judgement
$\sploot{\mathcal{J}}{\mathcal{J}'}$ (``$\InputMode{\mathcal{J}}$ sploots into
$\OutputMode{\mathcal{J}'}$'') and making it evident on all inputs:

\begin{gather*}
  \infer{
    \sploot{\mathcal{J}}{[x]\mathcal{J}'(x)}:\type
  }{
    \mathcal{J}:\sortoj{\Psi,A} &
    \mathcal{J}': \sortexp\to\sortoj\Psi
  }\\[6pt]
  \infer{
    \sploot{
      \qisset{\bind{x}{A(x)}}
    }{
      [x]\; \qisset{A(x)}
    }
  }{
  }\\[6pt]
  \infer{
    \sploot{
      \qeqset{\bind{x}{A(x)}}{\bind{x}{B(x)}}
    }{
      [x]\; \qeqset{A(x)}{B(x)}
    }
  }{
  }\\[6pt]
  \infer{
    \sploot{
      \qmem{\bind{x}{M(x)}}{\bind{x}{A(x)}}
    }{
      [x]\; \qmem{M(x)}{A(x)}
    }
  }{
  }\\[6pt]
  \infer{
    \sploot{
      \qeqmem{\bind{x}{M(x)}}{\bind{x}{N(x)}}{\bind{x}{A(x)}}
    }{
      [x]\; \qeqmem{M(x)}{N(x)}{A(x)}
    }
  }{
  }
\end{gather*}

Next, we will do something similar in order to show how each of these
judgement(-codes) behaves extensionally, by introducing another algorithmic
judgement $\refract{\mathcal{J}}{\mathcal{J}^=}$ (``$\InputMode{\mathcal{J}}$
refracts into $\OutputMode{\mathcal{J}^=}$''). Intuitively, this judgement
describes how what happens when you substitute two purportedly equal
expressions in for the judgement-code's outermost variable.

\begin{gather*}
  \infer{
    \refract{\mathcal{J}}{[x][y]\mathcal{J}'(x)(y)}:\type
  }{
    \mathcal{J}:\sortoj{\Psi,A} &
    \mathcal{J}': \sortexp\to\sortexp\to\sortoj\Psi
  }\\[6pt]
  \infer{
    \refract{
      \qisset{\bind{x}{A(x)}}
    }{
      [x][y]\; \qeqset{A(x)}{A(y)}
    }
  }{
  }\\[6pt]
  \infer{
    \refract{
      \qeqset{\bind{x}{A(x)}}{\bind{x}{B(x)}}
    }{
      [x][y]\; \qeqset{A(x)}{B(y)}
    }
  }{
  }\\[6pt]
  \infer{
    \refract{
      \qmem{\bind{x}{M(x)}}{\bind{x}{A(x)}}
    }{
      [x][y]\; \qeqmem{M(x)}{M(y)}{A(x)}
    }
  }{
  }\\[6pt]
  \infer{
    \refract{
      \qeqmem{\bind{x}{M(x)}}{\bind{x}{N(x)}}{\bind{x}{A(x)}}
    }{
      [x][y]\; \qeqmem{M(x)}{N(y)}{A(x)}
    }
  }{
  }
\end{gather*}

@@

We finally have built up enough machinery to give an immediate and elegant
encoding to Martin-L\"of's functional sequent judgement $\mathcal{J}\ (\Psi)$;
we will notate it $\sequent{\Psi}{\mathcal{J}}$, or $\nece{\mathcal{J}}$ when
$\Psi$ is the empty telescope. First we will define it, and then we'll go back
and show some corresponding examples between our system and the informal rules
of MLTT~1979.

\begin{gather*}
  \infer{
    \sequent{\Psi}{\mathcal{J}}:\type
  }{
    \Psi:\sorttele &
    \mathcal{J}:\sortoj\Psi
  }
  \quad
  \infer{
    \nece{\mathcal{J}} \equiv \sequent{\cdot}{\mathcal{J}} : \type
  }{
  }
\end{gather*}

@@

\begin{gather*}
  \infer={
    \nece{\qisset{\nobind{A}}}
  }{
    \isset{A}
  }
  \quad
  \infer={
    \nece{\qeqset{\nobind{A}}{\nobind{B}}}
  }{
    \eqset{A}{B}
  }\\[6pt]
  \infer={
    \nece{\qmem{\nobind{M}}{\nobind{A}}}
  }{
    \mem{M}{A}
  }
  \quad
  \infer={
    \nece{\qeqmem{\nobind{M}}{\nobind{N}}{\nobind{A}}}
  }{
    \eqmem{M}{N}{A}
  }
\end{gather*}

@@

\begin{gather*}
  \infer{
    \sequent{\Psi,A}{\mathcal{J}}
  }{
    \begin{array}{c}
      \sploot{\mathcal{J}}{[x]\mathcal{J}'(x)}\\
      \refract{\mathcal{J}}{[x][y]\mathcal{J}^=(x)(y)}\\
      \lfpi{x,x'}{
        \weaken{\Psi}{x}{x'}
        \to\sequent{\Psi}{\qmem{x'}{A}}
        \to\sequent{\Psi}{\mathcal{J}'(x)}
      }\\
      \lfpi{x,x',y,y'}{
        \weaken{\Psi}{x}{x'}
        \to\weaken{\Psi}{y}{y'}
        \to\sequent{\Psi}{\qeqmem{x'}{y'}{A}}
        \to\sequent{\Psi}{\mathcal{J}^=(x)(y)}
      }\\
    \end{array}
  }
\end{gather*}

@@

\theorem{The standard application rule for the cartesian product of a family of
sets is admissible:
\[
  \infer[\typi_E]{
    \mem{\ap(M;N)}{B(N)}
  }{
    \begin{array}{c}
      \mathcal{D}\\
      \mem{M}{\thunk{\typi(A;B)}}\\
    \end{array} &
    \begin{array}{c}
      \mathcal{E}\\
      \mem{N}{A}
    \end{array}
  }
\]
}
\begin{proof}
  By induction on the derivation $\mathcal{D}$, which must have the following canonical form:
  \[
     \infer[\mathcal{D}]{
       \mem{M}{\thunk{\typi(A;B)}}
     }{
       \begin{array}{c}
         \mathcal{D}' \\
         \eval{M}{\lam E}
       \end{array} &
       \infer{
         \canmem{\lam E}{\typi(A;B)}
       }{
         \infer{
           \sequent{\cdot,A}{\qmem{\tacitbind{E}}{\tacitbind{B}}}
         }{
           \infer{
             \lfpi{x,x'}{
               \weaken{\cdot}{x}{x'}
               \to\sequent{\cdot}{\qmem{x'}{A}}
               \to\sequent{\cdot}{\qmem{\nobind{E(x)}}{\nobind{B(x)}}}
             }
           }{
             \begin{array}{c}
               \mathcal{D}''\\
               \lfpi{x}{\mem{x}{A} \to \mem{E(x)}{B(x)}}
             \end{array}
           } &
           \dots
         }
       }
     }
  \]

  Admissibility follows from $\mathcal{D}'$, $\mathcal{D}''$, $\mathcal{E}$;
  the derivation $\mathcal{D''}(N)(\mathcal{E})$ must have a canonical form as follows:
  \[
    \infer[\mathcal{D''}(N)(\mathcal{E})]{
      \mem{E(N)}{B(N)}
    }{
      \begin{array}{c}
        \mathcal{F}\\
        \eval{E(N)}{X}
      \end{array} &
      \begin{array}{c}
        \mathcal{F}'\\
        \eval{B(N)}{C}
      \end{array} &
      \begin{array}{c}
        \mathcal{F}''\\
        \canmem{X}{C}
      \end{array}
    }
  \]
  Whence we may construct the goal:
  \[
    \infer{
      \mem{\ap(M;N)}{B(N)}
    }{
      \infer{
        \eval{\ap(M;N)}{X}
      }{
        \begin{array}{c}
          \mathcal{D}'\\
          \eval{M}{\lam E}
        \end{array} &
        \begin{array}{c}
          \mathcal{F}\\
          \eval{E(N)}{X}
        \end{array}
      }&
      \begin{array}{c}
        \mathcal{F}'\\
        \eval{B(N)}{C}
      \end{array}&
      \begin{array}{c}
        \mathcal{F}''\\
        \canmem{X}{C}
      \end{array}
    }
  \]
\end{proof}

\theorem{The two standard elimination rules for the disjoint union of a family of sets are admissible:
\[
  \infer[\tysg_{E_1}]{
    \mem{\fst(M)}{A}
  }{
    \mem{M}{\thunk{\tysg(A;B)}} &
  }\quad
  \infer[\tysg_{E_2}]{
    \mem{\snd(M)}{B(\fst(M))}
  }{
    \mem{M}{\thunk{\tysg(A;B)}} &
  }
\]
}
\begin{proof}
\end{proof}

>>>>

\newcommand\weaken[3]{\mathsf{wk}_{\InputMode{#1}}\,\InputMode{#2}\leadsto\OutputMode{#3}}
\newcommand\weakenrulenil{\mathsf{wk}_\mathsf{nil}}
\newcommand\weakenrulesnoc{\mathsf{wk}_\mathsf{snoc}}
\newcommand\qisset[1]{#1\;`\mathit{set}}
\newcommand\qeqset[2]{#1=#2\;`\mathit{set}}
\newcommand\qmem[2]{#1\;`\!\!\in#2}
\newcommand\qeqmem[3]{#1=#2\;`\!\!\in#3}

@@

\newcommand\sequentruleset{\gg_\mathsf{set}}
\newcommand\sequentruleeqset{\gg_{\mathsf{set}_=}}
\newcommand\sequentrulemem{\gg_\in}
\newcommand\sequentruleeqmem{\gg_{\in_=}}
\newcommand\dname[1]{\mathcal{#1}}
\newcommand\dlabel[2]{\begin{array}{c}#1\\#2\end{array}}

@@

The next thing to do is to specify how each of these judgement(-codes) behaves
operationally (i.e. by substituting its outermost variable); we will do so by
extending the LF signature with a new algorithmic judgement
$\sploot{\dname{J}}{\dname{J}'}$ (``$\InputMode{\dname{J}}$ sploots into
$\OutputMode{\dname{J}'}$'') and making it evident on all inputs:

\begin{gather*}
  \infer{
    \sploot{\dname{J}}{[x]\dname{J}'(x)}:\type
  }{
    \dname{J}:\sortoj{\Psi,A} &
    \dname{J}': \sortexp\to\sortoj\Psi
  }\\[6pt]
  \infer{
    \sploot{
      \qisset{\bind{x}{A(x)}}
    }{
      [x]\; \qisset{A(x)}
    }
  }{
  }\\[6pt]
  \infer{
    \sploot{
      \qeqset{\bind{x}{A(x)}}{\bind{x}{B(x)}}
    }{
      [x]\; \qeqset{A(x)}{B(x)}
    }
  }{
  }\\[6pt]
  \infer{
    \sploot{
      \qmem{\bind{x}{M(x)}}{\bind{x}{A(x)}}
    }{
      [x]\; \qmem{M(x)}{A(x)}
    }
  }{
  }\\[6pt]
  \infer{
    \sploot{
      \qeqmem{\bind{x}{M(x)}}{\bind{x}{N(x)}}{\bind{x}{A(x)}}
    }{
      [x]\; \qeqmem{M(x)}{N(x)}{A(x)}
    }
  }{
  }
\end{gather*}

Next, we will do something similar in order to show how each of these
judgement(-codes) behaves extensionally, by introducing another algorithmic
judgement $\refract{\dname{J}}{\dname{J}^=}$ (``$\InputMode{\dname{J}}$
refracts into $\OutputMode{\dname{J}^=}$''). Intuitively, this judgement
describes how what happens when you substitute two purportedly equal
expressions in for the judgement-code's outermost variable.

\begin{gather*}
  \infer{
    \refract{\dname{J}}{[x][y]\dname{J}'(x)(y)}:\type
  }{
    \dname{J}:\sortoj{\Psi,A} &
    \dname{J}': \sortexp\to\sortexp\to\sortoj\Psi
  }\\[6pt]
  \infer{
    \refract{
      \qisset{\bind{x}{A(x)}}
    }{
      [x][y]\; \qeqset{A(x)}{A(y)}
    }
  }{
  }\\[6pt]
  \infer{
    \refract{
      \qeqset{\bind{x}{A(x)}}{\bind{x}{B(x)}}
    }{
      [x][y]\; \qeqset{A(x)}{B(y)}
    }
  }{
  }\\[6pt]
  \infer{
    \refract{
      \qmem{\bind{x}{M(x)}}{\bind{x}{A(x)}}
    }{
      [x][y]\; \qeqmem{M(x)}{M(y)}{A(x)}
    }
  }{
  }\\[6pt]
  \infer{
    \refract{
      \qeqmem{\bind{x}{M(x)}}{\bind{x}{N(x)}}{\bind{x}{A(x)}}
    }{
      [x][y]\; \qeqmem{M(x)}{N(y)}{A(x)}
    }
  }{
  }
\end{gather*}

@@

We finally have built up enough machinery to give an immediate and elegant
encoding to Martin-L\"of's functional sequent judgement $\dname{J}\ (\Psi)$;
we will notate it $\sequent{\Psi}{\dname{J}}$, or $\nece{\dname{J}}$ when
$\Psi$ is the empty telescope. First we will define it, and then we'll go back
and show some corresponding examples between our system and the informal rules
of MLTT~1979.

\begin{gather*}
  \infer{
    \sequent{\Psi}{\dname{J}}:\type
  }{
    \Psi:\sorttele &
    \dname{J}:\sortoj\Psi
  }
  \quad
  \infer{
    \nece{\dname{J}} \equiv \sequent{\cdot}{\dname{J}} : \type
  }{
  }
\end{gather*}

@@

\begin{gather*}
  \infer=[\sequentruleset]{
    \nece{\qisset{\nobind{A}}}
  }{
    \isset{A}
  }
  \quad
  \infer=[\sequentruleeqset]{
    \nece{\qeqset{\nobind{A}}{\nobind{B}}}
  }{
    \eqset{A}{B}
  }\\[6pt]
  \infer=[\sequentrulemem]{
    \nece{\qmem{\nobind{M}}{\nobind{A}}}
  }{
    \mem{M}{A}
  }
  \quad
  \infer=[\sequentruleeqmem]{
    \nece{\qeqmem{\nobind{M}}{\nobind{N}}{\nobind{A}}}
  }{
    \eqmem{M}{N}{A}
  }
\end{gather*}

@@

\begin{gather*}
  \infer{
    \sequent{\Psi,A}{\dname{J}}
  }{
    \begin{array}{c}
      \sploot{\dname{J}}{[x]\dname{J}'(x)}\\
      \refract{\dname{J}}{[x][y]\dname{J}^=(x)(y)}\\
      \lfpi{x,x'}{
        \weaken{\Psi}{x}{x'}
        \to\sequent{\Psi}{\qmem{x'}{A}}
        \to\sequent{\Psi}{\dname{J}'(x)}
      }\\
      \lfpi{x,x',y,y'}{
        \weaken{\Psi}{x}{x'}
        \to\weaken{\Psi}{y}{y'}
        \to\sequent{\Psi}{\qeqmem{x'}{y'}{A}}
        \to\sequent{\Psi}{\dname{J}^=(x)(y)}
      }\\
    \end{array}
  }
\end{gather*}

@@

\theorem{The standard application rule for the cartesian product of a family of
sets is admissible:
\[
  \infer[\typi_E]{
    \mem{\ap(M;N)}{B(N)}
  }{
    \dlabel{\dname D}{
      \mem{M}{\thunk{\typi(A;B)}}
    } &
   \dlabel{\dname E}{
      \mem{N}{A}
    }
  }
\]
}
\begin{proof}
  By induction on the derivation $\dname{D}$, which must have the following canonical form:
  \[
     \infer[\dname{D}]{
       \mem{M}{\thunk{\typi(A;B)}}
     }{
       \dlabel{\dname D}{\eval{M}{\lam E}} &
       \infer{
         \canmem{\lam E}{\typi(A;B)}
       }{
         \infer{
           \sequent{\cdot,A}{\qmem{\tacitbind{E}}{\tacitbind{B}}}
         }{
           \dlabel{\dname D''}{
             \lfpi{x,x'}{
               \weaken{\cdot}{x}{x'}
               \to\sequent{\cdot}{\qmem{x'}{A}}
               \to\sequent{\cdot}{\qmem{\nobind{E(x)}}{\nobind{B(x)}}}
             }
           } &
           \dlabel{}{\dots}
         }
       }
     }
  \]

  Admissibility follows from $\dname{D}'$, $\dname{D}''$, $\dname{E}$;
  let $\dname{F} \equiv
  \dname{D}''(N,\nobind{N},\weakenrulenil,\sequentrulemem\dname{E})$, which
  must be a canonical derivation as follows:
  \[
    \infer[\dname{F}]{
      \sequent{\cdot}{\qmem{\nobind{E(N)}}{\nobind{B(N)}}}
    }{
      \infer{
        \mem{E(N)}{B(N)}
      }{
        \dlabel{\dname F'}{\eval{E(N)}{X}} &
        \dlabel{\dname F''}{\eval{B(N)}{C}} &
        \dlabel{\dname F'''}{\canmem{X}{C}}
      }
    }
  \]
  Whence we may construct the goal:
  \medskip

  \framebox{
    \begin{minipage}{0.9\textwidth}
      \[
        \infer{
          \mem{\ap(M;N)}{B(N)}
        }{
          \infer{
            \eval{\ap(M;N)}{X}
          }{
            \dlabel{\dname D'}{\eval{M}{\lam E}} &
            \dlabel{\dname F'}{\eval{E(N)}{X}}
          }&
          \dlabel{\dname F''}{\eval{B(N)}{C}} &
          \dlabel{\dname F'''}{\canmem{X}{C}}
        }
      \]
    \end{minipage}
  }
\end{proof}

\theorem{The two standard elimination rules for the disjoint union of a family of sets are admissible:
\[
  \infer[\tysg_{E_1}]{
    \mem{\fst(M)}{A}
  }{
    \dlabel{\dname D}{\mem{M}{\thunk{\tysg(A;B)}}}
  }\quad
  \infer[\tysg_{E_2}]{
    \mem{\snd(M)}{B(\fst(M))}
  }{
    \dlabel{\dname D}{\mem{M}{\thunk{\tysg(A;B)}}}
  }
\]
}
\begin{proof}

  The derivations of the elimination rules are constructed by induction on
  $\dname{D}$, which must be a canonical derivation of the following form:
  \[
    \infer[\dname{D}]{
      \mem{M}{\thunk{\tysg(A;B)}}
    }{
      \dlabel{\dname D_1}{
        \eval{M}{\pair(M_1,M_2)}
      } &
      \infer{
        \canmem{\pair(M_1,M_2)}{\tysg(A;B)}
      }{
        \infer{
          \mem{M_1}{A}
        }{
          \dlabel{\dname D_2}{\eval{M_1}{X}} &
          \dlabel{\dname D_3}{\eval{A}{A'}} &
          \dlabel{\dname D_4}{\canmem{X}{A'}}
        }&
        \infer[\dname D_5]{
          \mem{M_2}{B(M_1)}
        }{
          \dlabel{\dname D_6}{\eval{M_2}{Y}} &
          \dlabel{\dname D_7}{\eval{B(M_1)}{C}} &
          \dlabel{\dname D_8}{\canmem{Y}{C}}
        }
      }
    }
  \]

  The two elimination rules are derived as follows:
  \medskip

  \framebox{
    \begin{minipage}{0.9\textwidth}
      \begin{gather*}
        \infer[\tysg_{E_1}]{
          \mem{\fst(M)}{A}
        }{
          \infer{
            \eval{\fst(M)}{X}
          }{
            \dlabel{\dname D_1}{\eval{M}{\pair(M_1,M_2)}} &
            \dlabel{\dname D_2}{\eval{M_1}{X}}
          } &
          \dlabel{\dname D_3}{\eval{A}{A'}} &
          \dlabel{\dname D_4}{\canmem{X}{A'}}
        }\\[12pt]
        \infer[\tysg_{E_2}]{
          \mem{\snd(M)}{B(\fst(M))}
        }{
          \infer{
            \eval{\snd(M)}{Y}
          }{
            \dlabel{\dname D_1}{\eval{M}{\pair(M_1,M_2)}} &
            \dlabel{\dname D_6}{\eval{M_2}{Y}}
          }&
          \dlabel{\dname D_7}{\eval{B(\fst(M))}{C}} & % TODO: A little fishy
          \dlabel{\dname D_8}{\canmem{Y}{C}}
        }
      \end{gather*}
    \end{minipage}
  }
\end{proof}
